---
title: "Chapter 3"
author: "Anita Kurm"
date: "February 17, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#wd
setwd("C:/Users/JARVIS/Desktop/Uni/Semester 4/Computational modelling/Statistical-Rethinking")

#libraries
pacman::p_load(readr,groupdata2,ggplot2,tidyverse,data.table, rethinking)
```

Bayes' Theorem on a vampire example: 
  probability of one being a vampire given positive test for vampirism equals (Probability of a positive result given that the one is actually a vampire multiplied by probability of encountering a vampire in the entire population) divided by the average probability of a positive result  

    the average probability of a positive test result is calculated by:
      (probabilty of a positive result given that one is a vampire multiplied by probability of encountering a vampire in the population) + (probability of a positive result given that one is mortal multiplied by the probability of encountering mortals in the population)
      i.e. probability of true positives + probability of false positives

                       Pr(positive|vampire)*Pr(vampire)    Pr(positive|vampire)*Pr(vampire)
Pr(vampire|positive) = -------------------------------- = -------------------------------- 
                               Pr(positive)               Pr(positive|vampire)*Pr(vampire)+ 
                                                          Pr(positive|mortal)*Pr(mortal)
                                                          
                                                          Pr(mortal)= 1-Pr(vampire) duh... 
                                                          
```{r}
PrPV<-0.95 #test correctly detects vampirism 95% of the time
PrPM<-0.01 #false positives, 1% of the time it diagnoses normal people with vampirism
PrV<-0.001 #vampires are rare,only 0.001 of population
PrP<-PrPV*PrV+PrPM*(1-PrV)

(PrVP<-PrPV*PrV/PrP)
```

```{r}
#Performing calculation
```

