---
title: "Chapter 4"
author: "Anita Kurm"
date: "March 19, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#wd
setwd("C:/Users/JARVIS/Desktop/Uni/Semester 4/Computational modelling/Statistical-Rethinking")

#libraries
pacman::p_load(readr,groupdata2,ggplot2,tidyverse,data.table, rethinking)
```

Linear models
Linear regression as a Bayesian procedure: under a probability interpretation, linear regrssion uses a Gaussian distribution to describe model's uncertainty about some measurement of interest

  4.1. Why normal distributions are normal
    4.1.1. by addition
    football pitch experiment:
```{r}
#generate 16 random numbers per person representing individual steps of a random distance between 0 and 1 yard in either of the disrections <- runif(16,-1,1)
#add steps together to get the position after all 16 steps <- sum()
#replicate the procedure for 1000 people <- replicate(1000,...)
pos<- replicate(1000, sum(runif(16,-1,1)))

#plot the distribution
hist(pos)
plot(density(pos))
```
      we can see how normality emerges out of randomness
      -> random values "cancel each other out", i.e. a large positive fluctuation will cancel a large negative one -> convergence (more values in a sum, more values cancel each other out, so the sum tends to end up at zero relative to the mean.)

    4.1.2 by multiplication
    another way to get the normal distribution..
    loci and their growth effects multiply:
```{r}
#sample a random growth rate: sample 12 random numbers between 1.0 (no growth) and 1.1 (10% increase), representing a proportional increase in growth
prod(1+runif(12,0,0.1)) #the product of these 12 numbers is computed 

#generate 10000 random products
growth<- replicate(10000, prod(1+runif(12,0,0.1)))

#plot
dens(growth, norm.comp = T) #overlays normal distribution
```
      small effects that multiply together are approx additive  -> tend to stabilize on Gaussian distribution
```{r}
big<- replicate(10000, prod(1+runif(12,0,1)))
avg<- replicate(10000, prod(1+runif(12,0,0.4)))
small<- replicate(10000, prod(1+runif(12,0,0.01)))

#plot
dens(big, norm.comp = T)
mtext("BIG EFFECTS")
dens(avg, norm.comp = T)
mtext("BIG EFFECTS")
dens(small, norm.comp = T)
mtext("SMALL EFFECTS")
```
      
    With small effects the range of casual forces that tend towards Gaussian distributions extends beyond purely additice interactions
 
    4.1.3 By log-multiplication
    Large deviates (big effects) multiplied together do not produce Gaussian distributions, but they tend to produce Gaussian distributions on the log scale (so do the small deviates)
```{r}
log.big<- replicate(10000, log(prod(1+runif(12,0,1))))
dens(log.big, norm.comp = T)
mtext("BIG EFFECTS ON THE LOG SCALE")

log.small<- replicate(10000, log(prod(1+runif(12,0,0.01))))
dens(log.small, norm.comp = T)
mtext("SMALL EFFECTS ON THE LOG SCALE")
```
    Adding logs is equivalent to multiplying the original numbers ->
        log scales are used to measure sound and earthquakes and information (see chapter 6)

    4.1.4 Using Gaussian distributions
    - a member of exponential familyof fundamental natural distributions
    as a skeleton for hypotheses, building up models of measurements as aggregations of normal distributions. Justified by:
    
        4.1.4.1. Ontological justification
        it's a widespread (but not universal) pattern: measurement errors, variations in growth, velocities of molecules tend towards normal distribution <- cause these processes add together fluctuations(variations) -> results in a distribution of sums that have lost all info aside from mean and spread
        -> gaussian based models cannot reliably identify micro-process, still useful tho
        
         4.1.4.2. Epistemological justification
         it represents a particular state of ignorance: gaussian says only mean and variance of the distribution of measures
         a measure has finite variance <- it's least surprising and least informative assumption <- gaussian does not add any new assumptions <-gaussian is the most consistent with the model's assumptions
         this justification is premised on Inofrmation theory and maximum entropy
         
  4.2 A language for describing models
    1) Recognize outcome variable(s) - a set of measurements that we hope to predict or understand
    2) For each outcome variable, we define a likelihood distribution that defines the plausibility of individual observations <- Gaussian distribution in linear regression
    3) Recognize predictor variables - a set of other measurements we hope to use to predict or understand the outcome
    4) Relate the exact shape (location, variance etc) of the likelihood distribution to the predictor variables <- we are forced to name and define all of the parameters of the model 
    5) Choose priors for all of the parameters in the mode -> priors define the inital information state of the model, before seeing the data
    
    Models as mappings of one set of variables through a probability distribution onto another set of variables 
      4.2.1 Re-describing the globe tossing model
       w ~ Binomial(n,p)   <- the count w is distributed binomially with sample size n and probability p -> binom.distr. assumptions: each sample is independent of the others, sample points are independent of one another
       p ~ Uniform (0,1)   <- the prior p is assumed to be uniform between zero and one
       
       ~ means stochatic relationship (just a mapping of a variable or parameter onto a distribution), where no single instance of the variable on the left is known with certainty -> the mapping is probabilistic (some values are more plausible than others)
       
  4.3. A Gaussian model of height
    regression as soon as we have a predictor c´variable in it
    We want our Bayesian machine to consider every possible distribution (each defined by a combination of mu and sigma) and rank them by posterior plausibility - a measure of the logical compatibility of each possible distribution with the data and model
    -> we'll get an estimate of plausibility of every combination in light of data - a posterior distribution of Gaussian distributions ranked by plausibility
      4.3.1. The data
```{r}
#load the dataframe
data(Howell1)
d<- Howell1

#inspect the structure of tha dataframe
str(d)

#access the vector - column cont. the heights
d$height

#filter out non adults
d2<- d[d$age >= 18,]

dens(d2$height)
```
      4.3.2. The model
      the general model: hi ~ Normal(mu,sigma)    
      
      <- h refers to the list of heights, and the subscript i means each individual element of this list (index) <- the model knows that each height measurement is defined by the same normal distribution, with mean mu and std dev sigma
      the values hi are independent and identically distributed (iid)
      
      To complete the model, we need priors:
      we need a prior Pr(mu,sigma), since both mu and sigma are the parameters to be estimated - this prior is the joint probability for all parameters
      Pr(mu, sigma)= Pr(mu)Pr(sigma)
      
      So our model:
      hi ~ Normal(mu,sigma)     [likelihood]
      mu ~ Normal(178,20)       [mu prior]
      sigma ~ Uniform(0,50)     [sigma prior] flat
       
```{r}
#plot your priors

#prior for the mean - a broad gaussian prior centered on 178 (cause author is that tall, duh) with 95% of probability between 178+-40
curve(dnorm(x,178,20), from=100, to=250)   #the average height is almost certainly between 140 and 220cm - a bit of info from the prior

#prior for the std dev - a flat prior, a uniform one, so std dev has positive probability between 0 and 50cm
curve(dunif(x,0,50), from=-10, to=60)    #a std dev of 50cm implies that 95% ofindividual heights lie within 100cm of the average height

#sample from the prior
sample_mu <- rnorm(1e4, 178, 20)
sample_sigma <-runif(1e4, 0, 50)
prior_h <- rnorm(1e4, sample_mu, sample_sigma)
dens(prior_h)


```

      We get the distribution of relative plausibilities of different heights before seeing the data
      